---
title: "S&DS 230 Final Project: Gun Ownership, Violent Crime, and Gun Control Policy in the United States"
date: "Due: 2019 December 08"
output:
  pdf_document: default
  html_notebook: default
---


## Author:  Megan Zhang

## Discussants:

Articles mentioned, with link included:

[CBS News on Mass Shootings This Year](https://www.cbsnews.com/news/mass-shootings-2019-more-mass-shootings-than-days-so-far-this-year/)

[CNS News on More Guns, Less Violence?](https://www.cnsnews.com/commentary/cnsnewscom-staff/more-guns-less-gun-violence-between-1993-and-2013)

[A Blog Article on the Link between Gun Ownership and Deaths](https://medium.com/handwaving-freakoutery/everybodys-lying-about-the-link-between-gun-ownership-and-homicide-1108ed400be5)

[The Pew Research Center on Gun Deaths in the U.S.](https://www.pewresearch.org/fact-tank/2019/08/16/what-the-data-says-about-gun-deaths-in-the-u-s/)

[Harvard Injury Control Research Center on Availibility of Guns and Homicides](https://www.hsph.harvard.edu/hicrc/firearms-research/guns-and-death/)


<!--  


This is a template for creating your final project report. It lays out the sections that should be in your write-up and describes a little about these sections. There is some flexibility to deviate from this structure, for example, interweaving more visualizations and analyses could work well. 

Your report should be between 5-8 pages long and should contain:

    1) Introduction: 
      a) what is question you are addressing 
      b) why is important 
      c) where you got the data
      d) description of any other analyses that been done on the data
      
    2) Visualizations of the data: one or more plots
    
    3) Analyses: models, hypothesis tests, confidence intervals and other inferential statistics that give insight into your question
    
    4) Conclusions: What you found, future directions, etc.
    
    5) Reflection (do be completed on Canvas)
       a) What went well 
       b) What did you struggle with
       c) What analyses did you do that you are not including, etc. 

Please make your report look good by paying attention to detail, using additional R Markdown features etc. 

If there is additional code or data you would like to include with your report, please create a github page and add a link to it in your report. Additionally, you can append the full code for your analysis in an appendix section at the end of the document, and then include only the most important pieces of code in the body of the report. For example, you can exclude details of data cleaning from the body of the report, but include anything of central to your analyses, and also any information about particular choices you made that might affect the results and why you made those choices (e.g., explain if data was excluded and why, etc.). 



--> 










<!-- There are some options that might help make your document look better.  
Feel free to add adddtional options here -->
```{r message=FALSE, warning=FALSE, tidy=TRUE, echo=FALSE}

    library(knitr)
    library(dplyr)
    library(magrittr)

library(gridExtra)
library(grid)
library(ggplot2)
library(lattice)
library(tidyr)
    
    # This makes sure the code is wrapped to fit when it creats a pdf
    opts_chunk$set(tidy.opts=list(width.cutoff=60))   
    
```


## Introduction 


<!--  

Write ~1-3 paragraphs describing:

1) What is problem you are addressing and why the problem you are addressing is interesting

2) Where you got the data from, including a link to the website where you got the data if applicable 

3) What other analyses have already been done with the data and possibly links to other analyses, also mention if you are using the data in another class or for another research project. 


--> 

|     In 2019 to date, according to the Gun Violence Archive, there have been more mass shootings than days in the year. (a mass shooting is defined as any incident in which at least four people are shot. [(Link to news article.)](https://www.cbsnews.com/news/mass-shootings-2019-more-mass-shootings-than-days-so-far-this-year/)). The debate on gun control policy in the United State has become increasingly relevant and heated, with sharp divides in opinion, and usually along party lines. It seems intuitive to some that controlling gun access will decrease rates of violent crime across the country. On the other hand, some claim that controlling gun access only leads to use of other weaponry for crime and that they are needed for self-defense. Interestingly, CNSNews shows [(in this article)](https://www.cnsnews.com/commentary/cnsnewscom-staff/more-guns-less-gun-violence-between-1993-and-2013) that from 1993 to 2013, gun homicide rates in the United States actually decreased, while the number of guns per person increased. However, sites such as the Pew Research Center choose to focus on the past decade as a period of rapid increase in gun deaths [(link)](https://www.pewresearch.org/fact-tank/2019/08/16/what-the-data-says-about-gun-deaths-in-the-u-s/) rather than focusing on the overall decrease since the 1990s. With all of these opinions in mind, I decided to analyze the relationship between gun ownership and various crime metrics, including homicides and violent crime rates. Secondly, I will be looking at the specific gun laws of each state and comparing these metrics in order to determine whether stricter laws really do have an impact on crime rates. 

|     The data I am  using is compiled from several different sources onto the world population review website. Statistics for Gun Deaths by State [(link)](http://worldpopulationreview.com/states/gun-deaths-by-state/ ) are originally from the CDC's records of Firearm Mortality by State. Rates are alll recorded in deaths per 100k people. Data for Gun Laws by State [(link)](http://worldpopulationreview.com/states/gun-laws-by-state/) is compiled by GunsToCarry. Gun Ownership by State [(link)](http://worldpopulationreview.com/states/gun-ownership-by-state/) is based on 2017 gun registration statistics from the Federal Bureau of Alcohol, Tobacco, Firearms, and Explosives. Homicide and Violent Crime Rate by State [(link)](http://worldpopulationreview.com/states/crime-rate-by-state/) is reported from the FBI's Uniform Crime Reports from 2017. 

|     It does seem that many analyses on gun violence data have already been performed, to varying results. One writer claims that there is minimal correlation between gun ownership and gun violence deaths [(link)](https://medium.com/handwaving-freakoutery/everybodys-lying-about-the-link-between-gun-ownership-and-homicide-1108ed400be5). These analyses by the Harvard Injury Control Research Center [(link)](https://www.hsph.harvard.edu/hicrc/firearms-research/guns-and-death/) seem to support the notion that the increased availability of guns was correlated with increased homicides. Although I cannot reach the same level of depth of analysis as these studies, my goal in this analysis is to identify possible correlations or lack thereof. 


## Results


### Data wrangling: Combinining cleaned data into one data frame

```{r, warning= FALSE}
lawdat <- read.csv("gun_laws_by_state.csv")
ownershipdat <- read.csv("gun_ownership_by_state.csv") %>% select(State, gunOwnership)
homicidedat <- read.csv("homicide_rate_by_state.csv") %>% 
  select(State, homicideRate2017, firearmDeathRate) 
violentcrime <- read.csv("violent_crime_by_state.csv")

#Joining the data using merge() and left_join
completedat <- merge(ownershipdat, homicidedat, by ="State") %>%
  left_join(violentcrime, by = "State") %>% 
  merge(lawdat, by = "State")
completedat <- completedat %>% rename(homicideRate = homicideRate2017,
                                      violentcrimeRate = violentcrime_rate)
#Converting to narrow form for visualization
completedat_narrow <- tidyr::pivot_longer(completedat, 
                  cols = c(firearmDeathRate, homicideRate, violentcrimeRate))
completedat_narrow <- completedat_narrow %>% rename(crimeType = name,
                                                    rate = value)
```


<!--  

Very briefly discuss how you got the data into shape for your analyses. You can include some code here, although extensive data cleaning code should be put on github or in an appendix at the end of the document. 

--> 

Not much data cleaning was required because the calculated statistics (crime rate per capita, percentage of adults with gun ownership) were already made available by the World Population Review and its sources. I downloaded four separate datasets relating to gun ownership, firearm deaths, crime rates by state, and gun laws by state, and edited them in Excel as .csv files to make sure the variable names were appropriate. Then, I joined them all to one dataset using the State variable as the key. I also created a narrow version of the dataset to be used for visualization. 


### Visualizing the data: Comparing crime metrics against Gun Ownership and Gun Laws using ggplot
All code is listed in the Appendix. 

```{r, fig.width = 6, fig.height = 3, fig.align = "center", echo = FALSE}
ggplot(completedat, 
       aes(x = gunOwnership, y = firearmDeathRate, col = factor(permitReqToPurchase))) +
  geom_point() +
  xlab("Gun Ownership (Percentage of Adults)") +
  ylab("Firearm Death Rate (per 100k people)") +
  ggtitle("Firearm Death Rate vs. Gun Ownership")
```
**Figure 1.** The above plot depicts the relationship between gun ownership (measured in percentage of adults who own a gun) and the Firearm Death Rate for all fifty states in the U.S. As we can see there does seem to be a positive correlation, which we will analyze further in the next section. I have also separated the points according to factor which is whether or not the state requires gun purchasers to have a permit before they may purchase a gun. I chose this as the factor because of all the variables relating to gun law which we were given, a permit to purchase is considered one of the first steps in the process of purchasing a gun, and would present the first barrier to access. Additionally, the other factors relating to law were less commonly present throughout the states. 

$\\$
```{r,echo = FALSE, fig.width = 10, fig.height = 4.5}
plot1 <- ggplot(completedat, aes(x = gunOwnership, y = firearmDeathRate)) +
  geom_point() +
  xlab("Gun Ownership (%)") +
  ylab("Firearm Death Rate (per 100k people)") +
  ggtitle("Firearm Death Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_smooth(method = "lm", se = FALSE) 
  
plot2 <- ggplot(completedat, aes(x = gunOwnership, y = homicideRate)) +
  geom_point() +
  xlab("Gun Ownership (%)") +
  ylab("Homicide Rate (per 100k people)") +
  ggtitle("Homicide Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_smooth(method = "lm", se = FALSE)

plot3 <- ggplot(completedat, aes(x = gunOwnership, y = violentcrimeRate)) +
  geom_point() +
  xlab("Gun Ownership (%)") +
  ylab("Violent Crime Rate (per 100k people)") +
  ggtitle("Violent Crime Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_smooth(method = "lm", se = FALSE)

grid.arrange(plot1, plot2, plot3, ncol=3, 
             top = textGrob('Various Crime Rates against Gun Ownership', 
                            gp = gpar(fontsize = 15)))
```
**Figure 2.** This figure shows side-by-side plots of different crime metrics against gun ownership: First, firearm death rate (which is not officially a measure of crime), homicide rate, and violent crime rate. I chose to depict these three separately instead of using a factor variable because the scales were quite different (notice violent crime rates are much higher than the other two). Fitting a linear regression to each plot seems to show positive correlation for all three metrics; however, looking at the plotted points there really does not seem to be a clear correlation between gun ownership and homicide or violent crime rates. 

```{r, fig.height = 4, fig.width = 9, fig.align = "center", echo = FALSE}
plot4 <- ggplot(completedat, aes(x = factor(permitReqToPurchase), y = firearmDeathRate)) + 
  geom_boxplot(fill = "skyblue") + 
  labs(title = "Boxplot of Firearm Death Rate", 
       x = "Permit Required to Purchase", y = "Firearm Death Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_jitter()

plot5 <- ggplot(completedat, aes(x = factor(permitReqToPurchase), y = homicideRate)) + 
  geom_boxplot(fill = "skyblue") + 
  labs(title = "Boxplot of Homicide Rate", 
       x = "Permit Required to Purchase", y = "Homicide Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_jitter()

plot6 <- ggplot(completedat, aes(x = factor(permitReqToPurchase), y = violentcrimeRate)) + 
  geom_boxplot(fill = "skyblue") + 
  labs(title = "Boxplot of Violent Crime Rate", 
       x = "Permit Required to Purchase", y = "Violent Crime Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_jitter()

grid.arrange(plot4, plot5, plot6, ncol=3, 
             top = textGrob('Crime Rates by Gun Law: Permit to Purchase', 
                            gp = gpar(fontsize = 15)))

```
**Figure 3.** This figure shows side-by-side boxplots of each crime/death metric, divided according to whether or not the state requires a permit to purchase firearms. 
$\\$


<!--  

Create one or more plots of your data. Describe the plot(s), what they shows, and why they are of interest to your analysis. Include the code to create these plots in the R chunk below. You can also discuss the plots after the code too. Finally, be sure to make plots as clear as possible (clear axis labeling, legends and captions) so that it is easy for the reader to quickly understand the central information being conveyed. 

--> 

<!--  

Possible additional discussion of the plots here. 

--> 

**Discussion** Unsurprisingly, we can see that states with more relaxed gun control laws have higher rates of gun ownership, shown by the factor separation in Figure 1; additionally,  these seem to be correlated with higher gun death rates. Figure 2 seems to show a positive correlation between gun ownership and each of the metrics used; however, it does appear that using linear regression may not be the best choice for analysis since the points are relatively scattered. From Figure 3, it does appear that states which require a permit to purchase a firearm have lower Firearm Death Rate, Homicide Rates, and Violent Crime Rates. These differences will be explored in the Analysis section. 



$\\$    
    







    

### Analysis: Linear Models, Hypothesis tests of means, Confidence Intervals for Correlation and Regression Slopes

<!--  

Build linear models, run hypothesis tests, create confidence intervals and/or run simulations to answer questions that are of interest. 


--> 
**1. Fitting Linear Models for the Three Metrics**

Model for firearm Death Rate vs Gun Ownership. The slope coefficient is extracted from each linear model. 
```{r message=FALSE, warning=FALSE, tidy=TRUE}
#Models a Linear Regression for firearm death rate vs. gun ownership
lm_1 <-lm(firearmDeathRate ~ gunOwnership, data = completedat)
#Extracts the slope coefficient 
summary(lm_1)$coefficients[2,]
```

Homicide Rate vs Gun Ownership
```{r message=FALSE, warning=FALSE, tidy=TRUE}
lm_2 <- lm(homicideRate ~ gunOwnership, data = completedat)
summary(lm_2)$coefficients[2,]
```

Violent Crime vs. Gun Ownership
```{r message=FALSE, warning=FALSE, tidy=TRUE}
lm_3 <- lm(violentcrimeRate ~ gunOwnership, data = completedat)
summary(lm_3)$coefficients[2,]
```
**Discussion.** Here I fitted linear regression to each of the three plots from Figure 2. We can see the resulting linear regression in blue on each graph. All three gave slope values greater than 0. If we take the following null and alternative hypothesis:
$$
H_0: \beta_1 = 0;\,  \\
H_A: \beta_1 \neq 0;\, \\
\alpha = 0.05 
$$

We find that all three models give positive slopes with p-value less than $\alpha$, signifying a positive relationship. For firearm deaths, for example, each additional percentage in gun ownership appears to increase firearm death rate by about 25 per 100k people. This increase is 5.14 per 100k for homicides, and 427 for violent crimes. 


*However*, only the first model gives us a remotely high **r-squared value** of **0.487**, while the other two are very low (**0.078** and **0.134**). These values were calculated and extracted from the lm() function. Fitting **polynomials** up to degree 5 does give us higher r-squared and adjusted r-squared values (not shown in the code chunk above); however, I believe using this method will result in overfitting and looking at the data plotted above, it does not seem reasonable that any other degree polynomial would be a good fit for the data. Next, we'll create confidence intervals for correlation using the bootstrap method to see if the variables are actually correlated. 

**2. Confidence Interval for Correlation**

We examine $\rho$, the correlation constant, where $\rho = 0$ signifies no correlation. We will create $95\%$ confidence intervals for correlation between each of the three variables and gun ownership using the bootstrap method. 
```{r}
#Creating correlation interval using the Bootstrap Distribution 
bootstrap_dist <- NULL
for (i in 1:10000){
  one_bootstrap_data_frame <- completedat[sample(1:50, 50, replace = TRUE), ]
  bootstrap_dist[i] <- cor(one_bootstrap_data_frame$firearmDeathRate,
  one_bootstrap_data_frame$gunOwnership)
}

corint <- quantile(bootstrap_dist, c(0.025, 0.975))
```
Confidence interval for $\rho$ for firearm Death Rates and gun ownership : [`r corint`]

Using the same code as above for the other two metrics (see appendix for full code) we arrive at:

```{r, echo = FALSE}
#Creating correlation interval using the Bootstrap Distribution 
set.seed(1)
bootstrap_dist <- NULL
for (i in 1:10000){
  one_bootstrap_data_frame <- completedat[sample(1:50, 50, replace = TRUE), ]
  bootstrap_dist[i] <- cor(one_bootstrap_data_frame$homicideRate,
  one_bootstrap_data_frame$gunOwnership)
}

corint <- quantile(bootstrap_dist, c(0.025, 0.975))
```
Confidence interval for $\rho_{Homicide}$: [`r corint`].

```{r, echo = FALSE}
#Creating correlation interval using the Bootstrap Distribution 
bootstrap_dist <- NULL
for (i in 1:10000){
  one_bootstrap_data_frame <- completedat[sample(1:50, 50, replace = TRUE), ]
  bootstrap_dist[i] <- cor(one_bootstrap_data_frame$violentcrimeRate,
  one_bootstrap_data_frame$gunOwnership)
}

corint <- quantile(bootstrap_dist, c(0.025, 0.975))
```
Confidence interval for $\rho_{violentcrime}$: [`r corint`].

As we can see, the confidence interval for correlation between homicide and gun ownership contains 0, suggesting these two variables may not actually be correlated. The interval for Violent crime and gun ownership does not contain zero and implies a positive correlation; however, the interval is very close to 0, which I believe warrants further analysis on the correlation between these two variables. The confidence interval for $\rho$ of firearm death rates and gun ownership was high and did not contain 0. 

**3. Confidence Interval for Regression Slopes**

This confidence interval is using the bootstrap method. Since the sample size is small, and the data does not seem to come from a normal distribution (the variation in states is quite large), the bootstrap seems to be a more accurate method of estimation. 


Bootstrap Confidence interval for Slope of Firearm death rate vs. Gun Ownership:
```{r}
#Using the bootstrap to create a confidence interval for slope
nrep <- 10000
n_cases <- dim(completedat)[1]
result_vec <- rep(0, nrep)
for (i in 1:nrep){
  boot_sample <- dplyr::sample_n(completedat, size = n_cases, replace = TRUE)
  boot_fit <- lm(firearmDeathRate ~ gunOwnership, data = boot_sample)
  result_vec[i] <- coef(boot_fit)[2]
}

slopeconfint <- quantile(result_vec, c(0.025, 0.975))
```
The confidence interval is `slopeconfint` = [`r slopeconfint`].

```{r, echo = FALSE}
#Using the bootstrap to create a confidence interval 
nrep <- 10000
n_cases <- dim(completedat)[1]
result_vec <- rep(0, nrep)
for (i in 1:nrep){
boot_sample <- dplyr::sample_n(completedat, size = n_cases, replace = TRUE)
boot_fit <- lm(homicideRate ~ gunOwnership, data = boot_sample)
result_vec[i] <- coef(boot_fit)[2]
}

lm2boot <- quantile(result_vec, c(0.025, 0.975))
```
Using the same code as above (listed in the appendix) I calculated confidence intervals for the slopes of the remaining two metrics. 

Bootstrap Confidence interval for Slope ($\beta_1$) of Homicide Rate vs. Gun Ownership: The interval is [`r lm2boot`].
```{r, echo = FALSE}
#Using the bootstrap to create a confidence interval 
nrep <- 10000
n_cases <- dim(completedat)[1]
result_vec <- rep(0, nrep)
for (i in 1:nrep){
  boot_sample <- dplyr::sample_n(completedat, size = n_cases, replace = TRUE)
  boot_fit <- lm(violentcrimeRate ~ gunOwnership, data = boot_sample)
  result_vec[i] <- coef(boot_fit)[2]
}

lm3boot <- quantile(result_vec, c(0.025, 0.975))
```
Bootstrap Confidence interval for Slope ($\beta_1$) of Violent Crime Rate vs. Gun Ownership (code omitted): The interval is [`r lm3boot`].

Notice that the confidence interval for the slope of Homicide Rate vs. Gun Ownership contains 0, which implies there is not a significant linear relationship between the two variables.

**4. Comparison of Means: T test and Permutation Test**

In this section we compare the means on our crime-related metrics based on different laws established by the state. The first is whether or not the state requires a permit to purchase firearms. We'll run a t-test first in order to compare results of different tests. 
```{r}
#Separating the dataset on whether a permit is required to purchase
permit_req <- completedat %>% filter(permitReqToPurchase == "true")
permit_notreq <- completedat %>% filter(permitReqToPurchase == "false")

t.test(permit_req$firearmDeathRate, 
       permit_notreq$firearmDeathRate, alternative = "less")$p.value
t.test(permit_req$violentcrimeRate, 
       permit_notreq$violentcrimeRate, alternative = "less")$p.value
t.test(permit_req$homicideRate, 
       permit_notreq$homicideRate, alternative = "less")$p.value
```
I used the standard t.test() function to find p-values for comparison of each crime metric dependent on whether or not a permit in that state is required to purchase a gun. I used a one-tailed test and found the p values were all less than $\alpha = 0.05$, showing there is a significant negative difference between the means (the mean crime rate for states which require a permit is less than for states not requiring one). However, looking at distributions of the data reveals the points are not very normal; additionally, our sample size is not large, so it seems there are some concerns in performing a t-test. I think a more accurate test in this case would be a permutation test, which is what we will explore next. 

The following is an example of a permutation test using the data for Homicide Death Rates. 
```{r, fig.width = 6.5, fig.height = 4}
obs_stat <- mean(permit_req$homicideRate) - mean(permit_notreq$homicideRate)
combined <- c(permit_req$homicideRate, permit_notreq$homicideRate)
null_dist <- c()
for (i in 1:100000){
  shuff_data <- sample(combined)
  shuff_req <- shuff_data[1:14]
  shuff_notreq <- shuff_data[15:50]
  null_dist[i] <- mean(shuff_req) - mean(shuff_notreq)
}
hist(null_dist, main = "Histogram of Null Distribution Difference in Homicide Deaths",
  xlab = "Mean Difference",
  ylab = "Frequency",
  nclass = 100)
abline(v = obs_stat, col = "red")

#p-value for the observed Difference in mean
(p_val <- sum(null_dist <= obs_stat)/100000)
```

I used the same method to calculate a p-value for the observed difference in means for the other two crime metrics (Histograms omitted, and code included in the appendix). 

P-value for difference in means observed for Violent Crime Rates:
```{r, fig.width = 6.5, fig.height = 4, echo = FALSE}
obs_stat <- mean(permit_req$violentcrimeRate) - mean(permit_notreq$violentcrimeRate)
combined <- c(permit_req$violentcrimeRate, permit_notreq$violentcrimeRate)
null_dist <- c()
for (i in 1:100000){
  shuff_data <- sample(combined)
  shuff_req <- shuff_data[1:14]
  shuff_notreq <- shuff_data[15:50]
  null_dist[i] <- mean(shuff_req) - mean(shuff_notreq)
}

(p_val <- sum(null_dist <= obs_stat)/100000)
```

P-value for difference in means observed for Firearm Death Rates:
```{r, fig.width = 6.5, fig.height = 4, echo = FALSE}
obs_stat <- mean(permit_req$firearmDeathRate) - mean(permit_notreq$firearmDeathRate)
combined <- c(permit_req$firearmDeathRate, permit_notreq$firearmDeathRate)
null_dist <- c()
for (i in 1:100000){
  shuff_data <- sample(combined)
  shuff_req <- shuff_data[1:14]
  shuff_notreq <- shuff_data[15:50]
  null_dist[i] <- mean(shuff_req) - mean(shuff_notreq)
}

(p_val <- sum(null_dist <= obs_stat)/100000)
```
$\\$


    
    
    
    
    
    
\newpage

## Conclusion 



<!--  



1-2 paragraphs summarizing what you found, how the findings address your question of interest, and possible future directions. Please make sure describe your conclusions in an intuitive way, and make sure that your argument is strong and backed by solid evidence from your data. 



-->

|   In this analysis we were able to reach several interesting results. Firstly, there is positive correlation and linear relationship between gun ownership and firearm deaths. It seems fairly intuitive that the rate of gun deaths would increase as gun ownership also increases, so this result is not surprising to me. It seems obvious that if there are more guns available, there will be more deaths which result from guns, whether by murder or by accident. However, political discussions diverge on how useful this statistic actually is. According to the Pew Research Center, sixty percent of gun deaths in 2017 were suicides while 37$\%$ were murders. Some argue that the fact that most gun deaths are suicides means firearms are not linked to higher murder rates and thus additional regulation is not warranted. To examine this I looked at other metrics of crime and death such as homicide rates and violent crime rates, which I believe are more accurate metrics of how dangerous a city or place is. 

|   Surprisingly, I found that there does not seem to be statistically significant correlation between homicide rates and gun ownership, nor did there seem to be a significant linear relationship between the two variables. Additionally, while there seems to be a positive linear relationship between violent crime and gun ownership, the correlation between the two variables is low. However, one result which was consistent across all three metrics was that when looking at gun laws for each state, there was a statistically significant negative difference between crime rates and deaths in states that required permits as opposed to those that did not. It seems that on average, rates of violent crime, homicides, and firearm deaths are lower in states that require a permit to purchase firearms. I believe further study on this subject will be required in order to better understand the relationship between gun ownership, crime, and depths. The dataset I used for this project was very limited, in that it only contained one year's worth of data for only fifty states, which had much variation. Perhaps examining the same statistics over the past ten years would provide more comprehensive results; or, perhaps using the same variables but examining on a smaller scale such as cities or neighborhoods would be more helpful. Unfortunately, I tried but was unsucessful at finding such statistics. 









$\\$






## Reflection


<!--  


Reflection  

Write one paragraph describing what went well with this project and what was more difficult. Also describe any additional things you tried that you did not end up including in this write-up, and approximately how much time you spend working the project. 

Finally, please go to Canvas and answer a few questions related to how this project went: https://yale.instructure.com/courses/51220/quizzes/20279



-->

|   This project probably took me about 15 hours total. However, I spent at least 5 hours searching for a data set at the beginning of this project. I initially planned to do an analysis on the factors (e.g. demographic factors such as race, education, etc.) that influence opinions of American voters on gun control. However, this proved difficult because the dataset I was using consisted of mostly categorical variables, and it was difficult to use the methods we learned this semester to perform my analysis. When doing this project, I initially wanted to also analyze other gun laws, such as states which require registration or those that require background checks. I initially included a boxplot comparison of each crime metric according to whether or not the state required firarm registration, and intended to run tests to compare the means. There were several problems with this, however: Firstly, the number of states with each specific law was low, and the sample variation was too large to acheive much result. Also, not all laws were able to be described in simple true/false terms as the variables I used were. In the future I hope to learn more statistical methods so I will be able to analyze data that is of interest to me despite their limitations. 


$\\$


\newpage

## Appendix

###Data Visualization
```
#Figure 1
ggplot(completedat, 
       aes(x = gunOwnership, y = firearmDeathRate, col = factor(permitReqToPurchase))) +
  geom_point() +
  xlab("Gun Ownership (Percentage of Adults)") +
  ylab("Firearm Death Rate (per 100k people)") +
  ggtitle("Firearm Death Rate vs. Gun Ownership")
```

```
#Figure 2
plot1 <- ggplot(completedat, aes(x = gunOwnership, y = firearmDeathRate)) +
  geom_point() +
  xlab("Gun Ownership (%)") +
  ylab("Firearm Death Rate (per 100k people)") +
  ggtitle("Firearm Death Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_smooth(method = "lm", se = FALSE) 
  
plot2 <- ggplot(completedat, aes(x = gunOwnership, y = homicideRate)) +
  geom_point() +
  xlab("Gun Ownership (%)") +
  ylab("Homicide Rate (per 100k people)") +
  ggtitle("Homicide Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_smooth(method = "lm", se = FALSE)

plot3 <- ggplot(completedat, aes(x = gunOwnership, y = violentcrimeRate)) +
  geom_point() +
  xlab("Gun Ownership (%)") +
  ylab("Violent Crime Rate (per 100k people)") +
  ggtitle("Violent Crime Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_smooth(method = "lm", se = FALSE)

grid.arrange(plot1, plot2, plot3, ncol=3, 
             top = textGrob('Various Crime Rates against Gun Ownership', 
                            gp = gpar(fontsize = 15)))
```

```
#Figure 3
plot4 <- ggplot(completedat, aes(x = factor(permitReqToPurchase), y = firearmDeathRate)) + 
  geom_boxplot(fill = "skyblue") + 
  labs(title = "Boxplot of Firearm Death Rate", 
       x = "Permit Required to Purchase", y = "Firearm Death Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_jitter()

plot5 <- ggplot(completedat, aes(x = factor(permitReqToPurchase), y = homicideRate)) + 
  geom_boxplot(fill = "skyblue") + 
  labs(title = "Boxplot of Homicide Rate", 
       x = "Permit Required to Purchase", y = "Homicide Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_jitter()

plot6 <- ggplot(completedat, aes(x = factor(permitReqToPurchase), y = violentcrimeRate)) + 
  geom_boxplot(fill = "skyblue") + 
  labs(title = "Boxplot of Violent Crime Rate", 
       x = "Permit Required to Purchase", y = "Violent Crime Rate") +
  theme(plot.title = element_text(size=12)) +
  geom_jitter()

grid.arrange(plot4, plot5, plot6, ncol=3, 
             top = textGrob('Crime Rates by Gun Law: Permit to Purchase', 
                            gp = gpar(fontsize = 15)))

```
###Analysis 

```
#Omitted bootstrap distribution code for Correlation Interval 
#Creating correlation interval using the Bootstrap Distribution 
set.seed(1)
bootstrap_dist <- NULL
for (i in 1:10000){
  one_bootstrap_data_frame <- completedat[sample(1:50, 50, replace = TRUE), ]
  bootstrap_dist[i] <- cor(one_bootstrap_data_frame$homicideRate,
  one_bootstrap_data_frame$gunOwnership)
}

corint <- quantile(bootstrap_dist, c(0.025, 0.975))
```

```
#Creating correlation interval using the Bootstrap Distribution 
bootstrap_dist <- NULL
for (i in 1:10000){
  one_bootstrap_data_frame <- completedat[sample(1:50, 50, replace = TRUE), ]
  bootstrap_dist[i] <- cor(one_bootstrap_data_frame$violentcrimeRate,
  one_bootstrap_data_frame$gunOwnership)
}

corint <- quantile(bootstrap_dist, c(0.025, 0.975))
```

```
#Omitted Bootstrap code for difference in two means
obs_stat <- mean(permit_req$violentcrimeRate) - mean(permit_notreq$violentcrimeRate)
combined <- c(permit_req$violentcrimeRate, permit_notreq$violentcrimeRate)
null_dist <- c()
for (i in 1:100000){
  shuff_data <- sample(combined)
  shuff_req <- shuff_data[1:14]
  shuff_notreq <- shuff_data[15:50]
  null_dist[i] <- mean(shuff_req) - mean(shuff_notreq)
}

(p_val <- sum(null_dist <= obs_stat)/100000)
```
```
obs_stat <- mean(permit_req$firearmDeathRate) - mean(permit_notreq$firearmDeathRate)
combined <- c(permit_req$firearmDeathRate, permit_notreq$firearmDeathRate)
null_dist <- c()
for (i in 1:100000){
  shuff_data <- sample(combined)
  shuff_req <- shuff_data[1:14]
  shuff_notreq <- shuff_data[15:50]
  null_dist[i] <- mean(shuff_req) - mean(shuff_notreq)
}

(p_val <- sum(null_dist <= obs_stat)/100000)
```
<!--  


You can include a complete listing of your code here if you could not fit it into the body of the document. Make sure your code is well commmented and easy to create - i.e., use meaningful object names, separate your code into sections, describe what each section is doing, etc. 


-->



