---
title: "Class 18 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$





```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)


#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * Additional topics in simple linear regression: 
    * Confidence and prediction intervals, leverage, standarized and studentized residuals, Cook's distance
 
 * Multiple linear regression
 




$\\$






## Part 1: Additional statistics for evaluating simple linear regression fits


Let us use the data from the 2000 Florida election to look at additional metrics for evaluating simple linear regression fits. 






$\\$








#### Part 1.2: Loading the data and fitting the model


Let's start by loading the data and fitting a linear regression model predicting the number of 



```{r }


#download.file('https://raw.githubusercontent.com/emeyers/SDS230_F19/master/class_data/florida_vote_data_2000.Rda', 'florida_vote_data_2000.Rda')


load('florida_vote_data_2000.Rda')


florida_data <- dplyr::arrange(florida_data, Bush)

lm_fit <- lm(Buchanan ~ Bush, data = florida_data)




```







$\\$








#### Part 1.2: Confidence intervals and prediction intervals




```{r}


# confidence intervals for the betas
(CI_betas <- confint(lm_fit))


# confidence interval for the regression line mu_y
CI_regression_line <- predict(lm_fit, interval="confidence", level = 0.95)



# prediction interval for the regression line
prediction_interval <- predict(lm_fit, interval="predict")


# plot both confidence interval and the prediction interval
plot(Buchanan ~ Bush, data = florida_data, ylim = c(-1000, 3500))


# plot confidence interval
points(florida_data$Bush, CI_regression_line[, 1], col = "red", type = "l")
points(florida_data$Bush, CI_regression_line[, 2], col = "green", type = "l")
points(florida_data$Bush, CI_regression_line[, 3], col = "green", type = "l")


# plot prediction interval
points(florida_data$Bush, prediction_interval[, 2], col = "blue", type = "l")
points(florida_data$Bush, prediction_interval[, 3], col = "blue", type = "l")




```







$\\$







#### Part 1.3: Leverage and Cook's distance




```{r}

# get the h_i "hat values"
hat_vals <- hatvalues(lm_fit)

# plot the 20 largest hat values
names(hat_vals) <- florida_data$county_name
par(mar = c(4, 8, 4, 4))
barplot(tail(hat_vals, 20), las=2, xlab = "Leverage", horiz = TRUE)


# plot Cook's distances and plot them
plot(lm_fit, 4)


# get the names of the counties for 63 and 67
florida_data[c(63, 67), 2]


# plot residuals as a function of the hat values
plot(hat_vals, lm_fit$residuals, 
     xlab = "Leverage", 
     ylab = "Residuals")


# use the base R function to plot the standardized residuals with the hat values 
#  along with contours that contain constant Cook's distance values
plot(lm_fit, 5)


```






$\\$














