---
title: "Class 23 notes and code"
output:
  pdf_document: 
   fig_width: 4
   fig_height: 3
  html_document: default
---





$\\$





```{r setup, include=FALSE}
# install.packages("latex2exp")
library(latex2exp)
library(dplyr)
#options(scipen=999)
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```






$\\$






## Overview: Multiple linear regression continued



 * Plotting using grammar of graphics with ggplot
 * Bonus ggplot features
 * Joining data frames






$\\$








## Part 1: Visualization with ggplot


We can use the ggplot2 package which is part of the tidyverse to create much nicer looking graphics than using base R graphics. The ggplot2 library is modeled on Leland Wilkinson's "grammar of graphics" which tries to create graphics as a combination of basic visual elements. 


In the exercises below, we will use the motor trends cars data set (mtcars) that comes with base R installation, and also the gapminder data. 





$\\$





#### Part 1.1: Scatter plots


```{r}
# install.packages("ggplot2")
library(ggplot2)
# base R
plot(mtcars$wt, mtcars$mpg)
# ggplot - mapping in the geom
ggplot(data = mtcars) + 
  geom_point(mapping = aes(x = wt, y = mpg))
# ggplot - global mapping
 ggplot(data = mtcars, aes(x = wt, y = mpg)) + 
   geom_point()
# ggplot - shorter global mapping
ggplot(mtcars, aes(wt, mpg)) +
  geom_point()
```




$\\$





#### Part 1.2: Adding labels to plots


We can add labels to the plots using the xlab("label1") and ylab("label2") functions



```{r}
# ggplot - shorter global mapping
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() + 
  ylab("Miles per Gallon") + 
  xlab("Weight")
```





$\\$





#### Part 1.3: More aesthetic mappings


```{r}
# add color based on the transmission type (is automatic or not)
ggplot(mtcars, aes(x = wt, y = mpg, col = am)) +
  geom_point()
# it is better to treat the transmission type as a categorical variable
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(am))) +
  geom_point()
mtcars %>%
  mutate(am = as.factor(am)) %>%
  ggplot(aes(x = wt, y = mpg, col = am)) +
  geom_point()
# can also try mappping transmission type to shape or size 
mtcars %>%
  mutate(am = as.factor(am)) %>%     
  ggplot(aes(x = wt, y = mpg, size = am, shape = am, col = am)) +
  geom_point()
# mutating the names of a factor to be more meaningful 
mtcars %>%
  mutate(Transmission = ifelse(am == 0, "standard", "automatic")) %>%
  ggplot(aes(x = wt, y = mpg, col = Transmission)) +
  geom_point()
```



$\\$



#### Part 1.4: Attributes vs. Aesthetics


Setting an **aesthetic mapping** maps a variable to a **glyph property**

Setting an **attribute** set a glyph property to a **fixed value**



```{r}
# setting a aesthetic
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(am))) +
  geom_point()
# setting an attribute
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point(col = "red")
```




$\\$



#### Part 1.5: Facets


Beyond comparing variables based on aesthetics you can compare categorical variables by splitting a plot into subplots (called facets) using facet_wrap()




```{r}
# separate subplots for the two transmission types
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point() + 
  facet_wrap(~am)
# One can also do facets in two dimensions
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point() + 
  facet_grid(~vs ~am)
```




$\\$




#### Part 1.6: Overplotting


Sometimes points overlap making it hard to estimate the number of points at a particular range of values. 

We can control the transparency of points by changing their alpha values.




```{r}
library(gapminder)
# a lot of overplotting
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(alpha = .1)
# changing the transparency levels makes it a bit easir to see how many points are at a given x, y location
```





$\\$




#### Part 1.7: Changing scales


We can transform the scales of continuous data too


```{r}
# changing the scale on the x-axis
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(alpha = .5, col = "blue") + 
  scale_x_continuous(trans = "log10")
```






$\\$






#### Part 1.8: geoms


We can use different geoms to create other types of plots.



```{r}
# Let's create a plot that shows the GDP in the United States as a function of the year using the geom geom_line()
gapminder %>%
  filter(country == "United States") %>%
  ggplot(aes(x = year, y = gdpPercap)) + 
  geom_line() + 
  geom_point()
# Let's plot a histogram of the weights of cars
ggplot(mtcars, aes(x = wt)) + 
  geom_histogram(bins = 10)
# Let's create a boxplot of the weights of cars
ggplot(mtcars, aes(x = "", y = wt)) + 
  geom_boxplot()
# Let's create a side-by-side boxplot of the weights of cars depending on the number of cylinders the engine has
ggplot(mtcars, aes(x = factor(cyl), y = wt)) + 
  geom_boxplot(fill = "red")
```





$\\$





#### Part 1.9: geoms continued

Violin and Joy plots are other ways to view distributions of data



```{r}
# violin plot 
ggplot(mtcars, aes(x = factor(cyl), y = wt)) + 
  geom_violin(fill = "red") + 
  geom_jitter()
library("ggridges")
# joy plot
ggplot(mtcars, aes(y = factor(cyl), x = wt)) + 
  geom_density_ridges()
```


[Joy](https://f4.bcbits.com/img/a1100189446_10.jpg)





$\\$





#### Part 1.10: geoms continued continued


We can also have multiple geom layers on a single graph by using the + symbol
E.g   ggplot(.) + geom_type1() + geom_type2()



```{r}
# Create a scatter plot of miles per gallon as a function of weight and then add a smoothed line using geom_smooth()
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point() + 
  geom_smooth()
```





$\\$





#### Part 1.10: Themes


We can also use different types to change the appearance of our plot

```{r}
# Add theme_classic() to our plot
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point() + 
  theme_classic()
```




$\\$




#### Part 1.11: Example - plotting the IPED faculty salary data


```{r}
# source('../../../SDS230_F19/class_functions/class_functions.R')
# download_class_data("IPED_salaries_2016.rda")
load("IPED_salaries_2016.rda")
IPED_salaries %>%
  filter(CARNEGIE %in% c(15, 31)) %>%
  filter(rank_name %in% c("Lecturer", "Assistant", "Associate", "Full")) %>%
ggplot(aes(endowment, salary_tot, col = rank_name)) + 
  geom_point(alpha = .5) + 
  scale_x_continuous(trans = "log10") + 
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() + 
  xlab("Log10 Endowment") + 
  ylab("Salary ($)")
```

