---
title: "Class 11 notes and code"
output:
  html_document: default
  pdf_document: default
---





$\\$





```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)


options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * Correlation
 * Permutation tests for correlation
 * Theories of hypothesis tests
 * Parametric hypothesis tests



$\\$








## Part 1: of correlation


Let's review correlation and how we could create a permutation test to see if there is a statistically significant correlation between two quantitative variables. 





$\\$







## Part 2: Permutation tests for correlation





$\\$





##### Step 0: Plot the data


```{r cereal_plot}


# get the data
# source('class_functions.R')
# download_class_data("cereal.Rda")

# load the data
load("cereal.Rda")


# create a scatter plot of carbs as a function of calories
plot(cereal$Calories, cereal$Carbs)




```





$\\$







##### Step 1: State the null and alternative hypotheses 


$H_0: \rho = 0$
$H_A: \rho > 0$ 


$\alpha = $



$\\$








##### Step 2: Calculate the observed statistic




```{r sudoku_obs_stat}



# calculate the correlation statistic
(obs_stat <- cor(cereal$Calories, cereal$Carbs))
  
  

```






$\\$




##### Step 3: Create the null distribution




```{r sudoku_null_dist}


# one null distribution statistic 
cor(sample(cereal$Calories), cereal$Carbs)



# create the full null distribution 





```





$\\$







##### Step 4: Calculate a p-value


```{r sudoku_p_value}



# plot the null distribution again with a line at the observed statistic





# calculate the p-value



```


$\\$







##### Step 5: Make a decision







$\\$






## Homework 5: Was the 1969 Vietnam Draft completely random? 





$\\$